Uncovering underlying high-level musical content in the time domain \\ \small Leveraging self-supervised neural networks, inductive bias, and aural skills to learn deep audio embeddings with applications to boundary detection tasks

### 2023-08-21 discussion/discussion.tex
Updates from Overleaf

In this work, we introduced a method that leverages self-supervised deep neural networks to learn low-dimensional music latent representations with applications to music boundary detection. Building on existing approaches and architectures, we replaced traditional features with deep embeddings trained to represent high-level musical information to analyze its performance in music segmentation tasks.

@@

One of the main reasons why \cite{deepfeaturesegment, SalamonDeepSegmentation} outperform our method might be their task-tailored designs specifically oriented towards music segmentation analysis (MSA). In contrast, our method aims to be broader and more abstract, which may present a downside when evaluating specific tasks.

>>>>

In this work, we introduced a method that leverages self-supervised deep neural networks to learn low-dimensional music latent representations with applications to music boundary detection tasks. Building on existing approaches and architectures, we replaced traditional features with deep embeddings trained to represent high-level musical information to analyze its performance in music segmentation tasks.

@@

One of the main reasons why \cite{deepfeaturesegment, SalamonDeepSegmentation} might outperform our method might be their task-tailored and MSA-focused designs specifically oriented towards music segmentation. In contrast, our method aims to be broader and more abstract, which may present a downside when evaluating specific tasks.

### 2023-08-21 evaluation/evaluation.tex
Updates from Overleaf

\subsection{The GTZAN dataset}
The GTZAN Genre Collection is a widely-used dataset for music genre classification tasks. It comprises 1,000 audio tracks that are each 30 seconds long. These tracks are evenly distributed across ten genres: blues, classical, country, disco, hip hop, jazz, metal, pop, reggae, and rock, each containing precisely 100 songs. The audio files are stored as WAV files with a sample rate of 22050 Hz \cite{GTZAN}.

@@

Note: We used SALAMI's original release, 2011, featuring 1,359 tracks.

>>>>

\subsection{The GTZAN dataset}
The GTZAN Genre Collection is a widely-used dataset for music genre classification tasks. It comprises 1,000 audio tracks that are each 30 seconds long. These tracks are evenly distributed across ten genres: blues, classical, country, disco, hip hop, jazz, metal, pop, reggae, and rock, each containing precisely 100 songs. The audio files are stored as WAV files with a sample rate of 22,050 Hz \cite{GTZAN}.

@@

Note: We used SALAMI's original release from 2011, featuring 1,359 tracks.

### 2023-08-21 future work/future work.tex
Updates from Overleaf

\begin{itemize}
  \item Investigate the effect of different transformations and augmentation pipelines. Incorporating track stems and different takes of the same piece as natural, human-made augmentations could yield exciting results.
  \item Expand input data to include dB-scale Mel-spectrum magnitude and CQT of audio. This approach has been widely used in music-related tasks with CNNs \cite{Kim2020OneStrategies}. Though raw audio provides a rich representation, dB-scale Mel-spectrum offers a frequency-domain summarization that is not only grounded in psycho-acoustics but is also computationally efficient and hard to reproduce solely through data-driven methods. Therefore, this trade-off is worth exploring.
  \item Implement k-fold cross-validation to improve the robustness of the model's performance.
  \item Experiment with different hyperparameters. For instance, setting the kernel size to $0.005$ times the sample rate could match the Just Noticeable Difference (JND). As for the loss function margin, a starting point of 0.2 has proven effective, but different values should be explored to optimize performance on the validation set.
  \item Increase the size of the representation layer to 512 or 1024 dimensions.
  \item Apply easy triplet mining to improve the model's performance \cite{XuanImprovedMining}.
  \item Implement visual and auditory evaluations: 2D or 3D visualization of the latent space \ref{fig:manifold}, coupled with a Graphical User Interface (GUI) that enables playback for evaluation, can help assess the extent to which the model considers sonic attributes. It would also facilitate the understanding of the clustering of complex musical content. 
\end{itemize}

\input{figures/neural networks/manifold}

>>>>

\begin{itemize}
  \item Investigate the effect of different transformations and augmentation pipelines. Incorporating track stems and different takes of the same piece as natural, human-made augmentations could yield exciting results.
  \item Expand input data to include dB-scale Mel-spectrum magnitude and CQT of audio. This approach has been widely used in music-related tasks with CNNs \cite{deepfeaturesegment, Kim2020OneStrategies}. Though raw audio provides a rich representation, dB-scale Mel-spectrum offers a frequency-domain summarization that is not only grounded in psycho-acoustics but is also computationally efficient and hard to reproduce solely through data-driven methods. Therefore, this trade-off is worth exploring.
  \item Implement k-fold cross-validation to improve the robustness of the model's performance.
  \item Experiment with different hyperparameters. For instance, setting the kernel size to $0.005$ times the sample rate could match the Just Noticeable Difference (JND). As for the loss function margin, a starting point of 0.2 has proven effective, but different values should be explored to optimize performance on the validation set.
  \item Increase the size of the representation layer to 512 or 1024 dimensions.
  \item Apply easy triplet mining to improve the model's performance \cite{XuanImprovedMining}.
  \item Implement visual and auditory evaluations: 2D or 3D visualization of the latent space as displayed in Figure \ref{fig:manifold}, coupled with a Graphical User Interface (GUI) that enables playback for evaluation, can help assess the extent to which the model considers sonic attributes. It would also facilitate the understanding of the clustering of complex musical content. 
\end{itemize}

This represents a 'thorn in our side' that we intend to address in future research.

### 2023-08-21 introduction/Subsections/Assumptions.tex
Updates from Overleaf

This thesis posits that enduring, high-level musical elements persist regardless of sonic changes, mirroring the nature of the symbolic domain --sheet music--. A single waveform can correspond to a finite set of representations. Still, one sheet music excerpt can yield literally infinite interpretations.

>>>>

This thesis posits that enduring, high-level musical elements persist regardless of sonic changes, mirroring the nature of the symbolic domain. A single waveform can correspond to a finite set of representations. Still, one sheet music excerpt can yield infinite interpretations.

### 2023-08-21 introduction/Subsections/Introduction.tex
Updates from Overleaf

Beyond applications in music structure analysis, the evolution of neural networks has empowered the creation of latent representations \ref{fig:manifold} encapsulating crucial musical traits, technically known as representation ---or feature--- learning. This evolution has simplified the computational process for various task-specific elements, including but not limited to:

>>>>

Beyond applications in music structure analysis, the evolution of neural networks has empowered the creation of latent representations encapsulating crucial musical traits, technically known as representation ---or feature--- learning. Such spaces can be imploded into low dimensional and digestible representations, as shown in Figure \ref{fig:manifold}.

This evolution has simplified the computational process for various task-specific elements, including but not limited to:

### 2023-08-21 introduction/Subsections/Previous work.tex
Updates from Overleaf

Our approach is slightly different in two ways. First, we use physical time as input data rather than audio features. Second, our triplet selection pipeline is more abstract and designed for broader music understanding scope -yet still to be evaluated- rather than specifically tailored to music segmentation.

>>>>

Our approach is slightly different in two ways. First, we use physical time as input data rather than audio features. Second, our triplet selection pipeline is more abstract and designed for broader music understanding scope ---yet still to be evaluated--- rather than specifically tailored to music segmentation.

### 2023-08-21 methods/subsections/deep learning.tex
Updates from Overleaf

Deep learning (DL), a subset of machine learning, employs deep neural networks (DNNs) to decipher hierarchical data representations, allowing models with multiple processing layers to abstract data across numerous levels \cite{LeCun2015DeepLearning, Sarker2021DeepDirections}. DL models have significantly improved various domains due to their nonlinear modeling capabilities and scalability with large datasets.

They leverage the central algorithm, backpropagation, which calculates gradients of the loss function related to network parameters through forward and backward pass processes, thereby optimizing model performance.

@@

These computational models somewhat rooted in biological neural systems, utilize layers of artificial neurons for tasks such as pattern recognition, classification, and regression. They evolve by adjusting connections between neurons during learning, a concept initiated in the mid-20th century but modernized with Frank Rosenblatt's single-layer perceptron in 1957, capable of learning linearly separable patterns \cite{perceptron}.

@@

Another crucial component in the operation of DNNs is the activation function. This function is applied at every layer, transforming the linear combination of the input with the layer weights into an output that is passed onto the next layer. It introduces non-linearity into the network, enabling it to learn complex patterns and relationships.

>>>>

Deep learning (DL), a subset of machine learning (ML), employs deep neural networks (DNNs) to decipher hierarchical data representations, allowing models with multiple processing layers ---as shown in Figure \ref{fig: multilayer color perceptron}---to abstract data across numerous levels \cite{LeCun2015DeepLearning, Sarker2021DeepDirections}. DL models have significantly improved various domains due to their nonlinear modeling capabilities and scalability with large datasets.

They leverage the central algorithm, backpropagation, which calculates gradients of the loss function related to network parameters through forward and backward pass processes, thereby optimizing model performance. A high-level representation can be seen in Figure \ref{fig:error-backpropagation}.

@@

These computational models, somewhat rooted in biological neural systems, utilize layers of artificial neurons for tasks such as pattern recognition, classification, and regression. They evolve by adjusting connections between neurons during learning, a concept initiated in the mid-20th century but modernized with Frank Rosenblatt's single-layer perceptron in 1957, capable of learning linearly separable patterns \cite{perceptron}.

@@

Another crucial component in the operation of DNNs is the activation function. This function is applied at every layer, transforming the linear combination of the input with the layer weights into an output passed onto the next layer. It introduces non-linearity into the network, enabling it to learn complex patterns and relationships. See Figure \ref{fig:activation_function} for further visual understanding.

### 2023-08-21 methods/subsections/implementation.tex
Updates from Overleaf

In pattern recognition, models are designed to exhibit robustness against known invariances—transformations of input data—thereby ensuring consistent output. Even unknown invariances not explicitly considered in the model's design can be accommodated due to the model's inherent learning capacity.

Our implementation slightly modifies the CLMR method \cite{CLMR2021}, which learns valuable, discriminative music representations without explicit labels by contrasting positive augmentations of a musical piece against negative ones. CLMR falls under a branch of ML known as self-supervised learning (SSL) \cite{Balestriero2023ALearning}, where models learn autonomously from unlabeled data to create their own supervisory signals \cite{audioselfsupsurvey}. This method resembles how humans learn from observations and interactions, transforming unsupervised problems into supervised ones by auto-generating labels. SSL benefits include reduced dependence on labeled data, contributing to more robust and generalizable data representations.

@@

With only 2.4 million trainable parameters, this fully convolutional model reduces computational requirements and learns features at different scales through its multi-resolution architecture \ref{tab:samplecnntable}.

@@

The learning rate, set at 0.003, is a pivotal parameter dictating the step size at each iteration towards loss function minimization. It's a delicate balancing act— a high rate promises swift convergence with a risk of minimum overshoot, while a lower rate provides careful convergence but necessitates more iterations. Given that delicacy, we set it to a standard number broadly used in the literature.

@@

Previous research has utilized CNNs with various features such as Mel-Scaled Log-magnitude Spectograms (MLS), Self-Similarity Matrices (SSM), and Self-Similarity Lag Matrices (SSLM) as inputs \cite{Hernandez-Olivan2021MusicFeatures}. However, features derived from raw audio may lack interpretability in some scenarios \cite{Schindler2020DeepTutorial}, and raw audio presents unique advantages despite being highly computationally demanding. It ensures the preservation of the original signal, potentially uncovering novel insights, and allows for direct feature extraction via advanced DL models \cite{learning, verydeep}. Nevertheless, it comes with challenges, such as high dimensionality requiring substantial computational resources.

Time-domain processing naturally handles temporal patterns and sequences in data, thereby avoiding windowing artifacts. Although audio feature-based methods are effective for various audio-related machine learning tasks, their limitations lie in representing perceptual similarity. MLS, for example, captures frequency distribution over time. Yet, the complexity of human auditory perception—encompassing temporal patterns, phase relationships between frequencies, and higher-level musical structures means that musically similar sounds can have distinct spectrograms. This discrepancy implies that using spectrogram distance alone for measuring high-level music content may not always align with human perceptions \cite{Kim2020OneStrategies, Mesostructures2023}.

@@

The positive image in every triplet of input data must preserve its intelligible content when subjected to transformations, regardless of alterations in sonic qualities and processing artifacts. Maintaining the temporal structure and meaningfulness of the content allows it to present musical elements remarkably close to the original track.

While we have experimented with helpful audio augmentation tools such \cite{Spijkervet2021Spijkervet/torchaudio-augmentations:V1.0, Kharitonov2020DataDomain}, the specific requirements of our experiments required the development of our own transformation chain using \textit{torchaudio}'s \cite{Yang2021TorchAudio:Processing} implementation of \textit{SoX} \cite{sox}: given an anchor audio signal $A[n]$, we generate a positive signal $P[n]$ by applying a series of amplitude, time-domain, frequency-domain, modulation, reverberation, and nonlinear effects with additive noise on top of it.

@@

The positive signal $P[n]$ is generated by convoluting the various impulse responses of specific effects. The noise signals are included with a specific SNR on top of the effect chain.

Random parameter updates within hardcoded ranges generate unique audio $P[n]$ out of $A[n]$ for each run-through. Adding random white noise and varying SNR creates countless noisy waveform variations. Although possible combinations can be estimated by multiplying discrete parameter values, the presence of continuous parameters and randomness in noise generation effectively results in infinite unique audio versions.

@@

While some packages can be found in the MIR online community \cite{auraloss}, we wrote our own PyTorch \cite{Paszke2019PyTorch:Library} implementation for our experiments.

@@

The \textit{embeddiogram}, a deep audio feature representation, is derived by applying our pre-trained neural network to sliding windowed segments across the audio signal, generating a sequence of embedding vectors. These relatively low-dimensional vectors collectively form a two-dimensional description of the audio signal's musical content.

@@

\item \textbf{Normalize the \textit{embeddiogram \ref{fig:embeddiogram}}}: The \textit{embeddiogram} is normalized to have a minimum value of 0 and a maximum value of 1 per dimension. The normalization process is given by:
\begin{equation}
E'_{ij} = \frac{E_{ij} - \min(E)}{\max(E) - \min(E)}
\end{equation}
\end{enumerate}

\begin{figure}[ht]
  \centering
  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/355embeddiogramnormalized.png}
    \caption[Embeddiogram. Track 355 (SALAMI dataset).]{Embeddiogram. Track 355 (SALAMI dataset).}
    \label{fig:embeddiogram}
  \end{minipage}

  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/355novelty.png}
    \caption[Novelty curve + peaks. Track 355 (SALAMI dataset).]{Novelty curve and peak detection. Track 355 (SALAMI dataset).}
    \label{fig:novelty}
  \end{minipage}
\end{figure}

@@

These deep audio features can be a foundation for current state-of-the-art methods, which are expected to receive as input standard traditional audio features. Consequently, they can be processed and manipulated like conventional features as described by \cite{unsuperMSA} and displayed in \ref{fig:novelty}, \ref{fig:SSM}, \ref{fig:SSLM}, \ref{fig:Q}, and \ref{fig:TBSM}. Employing such embeddings in a traditional music segmentation algorithm can achieve state-of-the-art performance \cite{deepfeaturesegment}.

>>>>

In pattern recognition, models are designed to exhibit robustness against known invariances—transformations of input data ---thereby ensuring consistent output. Even unknown invariances not explicitly considered in the model's design can be accommodated due to the model's inherent learning capacity.

Our implementation slightly modifies the Contrastive Learning of Music Representations (CLMR) method \cite{CLMR2021}, which learns valuable, discriminative music representations without explicit labels by contrasting positive augmentations of a musical piece against negative ones. CLMR falls under a branch of ML known as self-supervised learning (SSL) \cite{Balestriero2023ALearning}, where models learn autonomously from unlabeled data to create their own supervisory signals \cite{audioselfsupsurvey}. This method resembles how humans learn from observations and interactions, transforming unsupervised problems into supervised ones by auto-generating labels. SSL benefits include reduced dependence on labeled data, contributing to more robust and generalizable data representations.

@@

With only 2.4 million trainable parameters, this fully convolutional model reduces computational requirements and learns features at different scales through its multi-resolution architecture. See layer specifications in Table \ref{tab:samplecnntable}.

@@

The learning rate, set at 0.003, is a pivotal parameter dictating the step size at each iteration towards loss function minimization. It's a delicate balancing act ---a high rate promises swift convergence with a risk of minimum overshoot, while a lower rate provides careful convergence but necessitates more iterations. Given that delicacy, we set it to a standard number broadly used in the literature.

@@

Previous research has utilized CNNs with various features such as Mel-Scaled Log-magnitude Spectograms (MLS), Self-Similarity Matrices (SSM), and Self-Similarity Lag Matrices (SSLM) as inputs \cite{Hernandez-Olivan2021MusicFeatures}. However, features derived from raw audio may lack interpretability in some scenarios \cite{Schindler2020DeepTutorial}, and raw audio presents unique advantages despite being highly computationally demanding. It ensures the preservation of the original signal, potentially uncovering novel insights, and allows for direct feature extraction via advanced DL models \cite{learning, verydeep}. Nevertheless, it comes with challenges, such as high ---the highest--- dimensionality requiring substantial computational resources.

Time-domain processing naturally handles temporal patterns and sequences in data, thereby avoiding windowing artifacts. Although audio feature-based methods are effective for various audio-related machine learning tasks, their limitations lie in representing perceptual similarity. MLS, for example, captures frequency distribution over time. Yet, the complexity of human auditory perception, encompassing temporal patterns, phase relationships between frequencies, and higher-level musical structures means that musically similar sounds can have distinct spectrograms. This discrepancy implies that using spectrogram distance alone for measuring high-level music content may not always align with human perceptions \cite{Kim2020OneStrategies, Mesostructures2023}.

@@

The positive waveform in every triplet of input data must preserve its intelligible content when subjected to transformations, regardless of alterations in sonic qualities and processing artifacts. Maintaining the temporal structure and meaningfulness of the content allows it to present musical elements remarkably close to the original track.

While we have experimented with helpful audio augmentation tools such as \cite{Spijkervet2021Spijkervet/torchaudio-augmentations:V1.0, Kharitonov2020DataDomain}, the specific requirements of our experiments required the development of our own transformation chain using \textit{torchaudio}'s \cite{Yang2021TorchAudio:Processing} implementation of \textit{SoX} \cite{sox}: given an anchor audio signal $A[n]$, we generate a positive signal $P[n]$ by applying a series of amplitude, time-domain, frequency-domain, modulation, reverberation, and nonlinear effects with additive noise on top of it.

@@

The positive signal $P[n]$ is generated by convoluting the various impulse responses of specific effects. The noise signals are included with a set SNR ratio within the above range on top of the effect chain.

Random parameter updates within hardcoded ranges generate unique audio $P[n]$ images out of $A[n]$ anchor for each run-through. Adding random white noise and varying SNR creates countless noisy waveform variations. Although possible combinations can be estimated by multiplying discrete parameter values, the presence of continuous parameters and randomness in noise generation effectively results in infinite unique audio versions.

@@

While some packages can be found in the MIR online community \cite{auraloss}, we wrote our own \textit{PyTorch} \cite{Paszke2019PyTorch:Library} implementation for the sake of our experiments.

@@

The \textit{embeddiogram}, a deep audio feature representation, is derived by applying our pre-trained neural network to sliding windowed segments across the audio signal, generating a sequence of embedding vectors. These relatively low-dimensional vectors collectively form a two-dimensional description of the audio signal's musical content. A visual representation can be seen in Figure \ref{fig:embeddiogram}.

@@

\item \textbf{Normalize the \textit{embeddiogram}}: The \textit{embeddiogram} is normalized to have a minimum value of 0 and a maximum value of 1 per dimension. The normalization process is given by:
\begin{equation}
E'_{ij} = \frac{E_{ij} - \min(E)}{\max(E) - \min(E)}
\end{equation}
\end{enumerate}

\begin{figure}[ht]
  \centering
  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/355embeddiogramnormalized.png}
    \caption[Embeddiogram | Track 355 (SALAMI dataset).]{Embeddiogram. Track 355 (SALAMI dataset).}
    \label{fig:embeddiogram}
  \end{minipage}

  \begin{minipage}[b]{1.0\linewidth}
    \centering
    \includegraphics[width=\linewidth]{figures/images/355novelty.png}
    \caption[Novelty curve and peaks | Track 355 (SALAMI dataset).]{Novelty curve and peak detection. Track 355 (SALAMI dataset).}
    \label{fig:novelty}
  \end{minipage}
\end{figure}

@@

These deep audio features can be a foundation for current state-of-the-art methods, which are expected to receive as input standard traditional audio features. Consequently, they can be processed and manipulated like conventional features as described by \cite{unsuperMSA} and displayed in Figures \ref{fig:novelty}, \ref{fig:SSM}, \ref{fig:SSLM}, \ref{fig:Q}, and \ref{fig:TBSM}. Employing such embeddings in a traditional music segmentation algorithm can achieve state-of-the-art performance \cite{deepfeaturesegment}.

### 2023-08-21 results/subsections/tables.tex
Updates from Overleaf

Table \ref{ta:results} provide a comparative analysis across four features using three segmentation algorithms. The results reveal that the performance of the algorithms varies depending on the feature type. For instance, the CQT feature exhibits the highest precision (0.570), recall (0.339), and f-measure (0.353) when processed with the Foote algorithm. Nevertheless, this table proves that our embeddiograms can achieve competitive performance compared to traditional handcrafted signal processing methods by learning only from unlabeled audio files.

The representation of these results is also provided in Figure \ref{fig:boxplotmetrics}, to understand the metrics' distribution for each feature visually. The high outliers suggest that there is significant variability in the performance of the model. While the model performs exceptionally well in some instances, it performs average in most cases.

>>>>

Table \ref{ta:results} provides a comparative analysis across four features using three segmentation algorithms. The results reveal that the performance of the algorithms varies depending on the feature type. For instance, the CQT feature exhibits the highest precision (0.570), recall (0.339), and f-measure (0.353) when processed with the Foote algorithm. Nevertheless, this table proves that our embeddiograms can achieve competitive performance compared to traditional handcrafted signal processing methods by learning only from unlabeled audio files.

These results are also represented in Figure \ref{fig:boxplotmetrics} to understand the metrics' distribution for each feature visually. The high outliers suggest that there is significant variability in the performance of the model. While the model performs exceptionally well in some instances, it performs average in most cases.
